How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-24 talk/dyla2007/talk.tex
start the "problems with C" part

% examples from my last beamer talk, shows the most important features
\begin{frame}
  \frametitle{Pyrolog}

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in Python –
  \item
    want to compile it to get interesting performance
  \item
    translation tool-chain part of the PyPy project
  \item
    Python itself is too dynamic to be translatable to other languages, need a subset
  \item
    RPython (``Restricted Python'') is a subset of Python translatable to other
    languages
  \item
    RPython is designed to be significantly faster than regular Python
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Python case}
  CPython (the reference implementation) is a straightforward, portable VM.

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, global lock \dots
  \item
    No dynamic compilation
  \end{itemize}
  \pause
  \begin{block}{
    Extensions:}
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, green threads)
    \item
      \alert{Psyco} (run-time specializing compiler)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{block}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Introduction}
  \begin{itemize}
  \item XXX intro, talk about scope
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Common Approaches to VM construction}
  \begin{block}{
    Using C directly (or C disguised as another language)}
    \begin{itemize}
    \item
      CPython
    \item
      Ruby
    \item
      Spidermonkey (Mozilla's JavaScript VM)
    \item
      but also: Squeak, Scheme48
    \end{itemize}
  \end{block}
  \begin{block}{
    Building on top of an existing OO VM}
    \begin{itemize}
    \item
      Jython, IronPython
    \item
      JRuby, IronRuby
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The Problems of C}
  When writing a VM in C it is hard to reconcile:
  \begin{itemize}
  \item
    flexibility, maintainability
  \item
    simplicity of the VM
  \item
    performance (needs dynamic compilation techniques)
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      \alert{CPython} is a very simple bytecode VM, performance not great
    \item
      \alert{Psyco} is a just-in-time-specializer, very complex, hard to
      maintain, but good performance
    \item
      \alert{Stackless} is a fork of CPython adding microthreads. It was never
      incorporated into CPython for complexity reasons
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Compilers are a bad encoding of Semantics}
  \begin{itemize}
  \item to reach good performance levels, dynamic compilation is often needed
  \item a dynamic compiler needs to encoding language semantics
  \item this encoding is often relatively obscure and hard to change
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Psyco is a dynamic compiler for Python
    \item
      synchronizing with CPython's rapid development is a lot of effort
    \item
      many of CPython's new features not supported well
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Fixing of Early Design Decisions}
  \begin{itemize}
  \item when starting a VM in C, many decisions need to be made immediately
  \item examples: memory management technique, threading model
  \item the decision is manifested throughout the VM source
  \item very hard to change later
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      CPython uses reference counting, increfs and decrefs everywhere
    \item
      CPython uses OS threads with one global lock, hard to change to
      lightweight threads or finer locking
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementation Proliferation}
  \begin{itemize}
  \item
    restrictions of the original implementation lead to re-implementations,
    forks
  \item
    all implementations need to be syncronized with language evolution
  \item
    lots of duplicate effort
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      several serious implementations: CPython, Stackless, Psyco, Jython,
      IronPython, PyPy
    \item
      the implementations have various grades of compliance
    \end{itemize}
  \end{block}
\end{frame}
